\section{Assault}

\begin{enumerate}
    \item The threat or use of force on another that causes that person to have a \textbf{reasonable apprehension of imminent harmful or offensive contact}.
    \item The Second Restatement does not require apprehension to be ``reasonable,'' but most courts do.
    \item Assault in torts is different than assaults in criminal law.
    \item Can damages be awarded if physical harm did not occur? \textbf{\emph{I de S et Ux v. W de S}}: Defendant tries to buy wine from the plaintiff. He beats on the door with a hatchet. When the plaintiff's wife asks him to stop, he tries to hit her with the hatchet (but did not hit her). The court ruled that the plaintiff was entitled to damages even though no physical harm was done.
    \item Can forward looking verbal threats suggest imminent harm? \textbf{\emph{Castro v. Local 1199, National Health \& Human Services Employees Union}}: Plaintiff has asthma, which prevented her from working in extremely hot or cold situations. After a disability leave, she attended a meeting where she didn't receive her usual assignment. She asked what was going on, and Frankel (another employee) replied, ``If I was you, I would take whatever they give me, because you could lose more than your job.'' When asked he was threatening her life, Frankel said, ``Take it any way you want.'' The court held that \textbf{verbal threats, without ``circumstances inducing a reasonable apprehension of bodily harm,'' do not constitute an assault}. Here, the threat was ``forward-looking'' and did not suggest imminent harm. The court granted the defendant's motion for summary judgment.
    \item More questions on assault:
    \begin{enumerate}
        \item Is it possible to rationalize the difference between the criminal and tort definitions of assault?
        \item Why prevent assault?
        \item Why must assault be imminent?
        \item Can words alone ever be enough to constitute assault? See \emph{Campbell v. Kansas State University}, where a university head said ``he felt like hitting his assisstant on the buttocks, after he had already slapped her on the buttocks,'' which the court held to be assault (Diamond 20).
    \end{enumerate}
\end{enumerate}
