\subsection{Cause in Fact}

\begin{enumerate}
    \item Plaintiff must show that the defendant's negligence was a cause in 
    fact of the harm.
    \item Traditionally: plaintiff must prove that the harm would not have 
    occurred \textbf{but for} the defendant's actions.
    \item If there are multiple causes of harm, each can be but-for causes as 
    long as the harm would not have occurred without it.
    \item If there are multiple causes of harm, but none alone is a but-for 
    cause, courts can use the \textbf{substantial factor test}. See 
    \emph{Northington} below.
    \item The biggest different between but-for test and substantial factor 
    test is that under the substantial factor test, it's much more likely to 
    go to a jury.
    \item \textbf{Proximate cause} removes liability when ``the connection 
    between the plaintiff's harm and defendant's liability is unforeseeable or 
    so attenuated that public policy precludes liability.''\footnote{Casebook 
    p. 206.}
    \item When two people are \textbf{``acting in concert''} (i.e., trying to 
    do the same thing), and one is the negligent actor, the court can hold 
    both parties liable.
    \item \emph{Summers v. Tice}, below: the California Supreme Court adopted 
    the \textbf{alternative liability test}: when one of two negligent 
    defendants probably caused a harm, and it has not been shown that it is 
    more likely than not that either caused it, then \textbf{each will be held 
    jointly and severally liable for the full amount of the harm}.
    \begin{enumerate}
        \item Restatement Second says the alternative liability test applies 
        when there are ``two or more'' defendants.
        \item Also think about the effect of Prop 51 on \emph{Summers v. 
        Tice}: not clear whether it applies to these joint tortfeasor cases or 
        not.
    \end{enumerate}
    \item \emph{Sindell}, below, established \textbf{market share liability}: 
    each defendant shall be held liable for the proportion of the judgment 
    represented by its share of the market unless it can demonstrate that it 
    did not manufacture the product that caused the plaintiffs' injuries. This 
    is the case in California, but not in New York.
    \item Toxic torts: what to do when there is no harm, but only enhanced 
    risk? One approach: award damages, but to a lesser amount, based on the 
    percent chance of the harm actually occurring. Seond approach (which is 
    the rule in California): the court will not award general damages, but it 
    will award damages for the costs of relevant medical surveillance.
\end{enumerate}

\subsubsection{Proving Cause in Fact: \emph{East Texas Theatres, Inc. v. Rutledge}}

To establish causation, the cause must have been the cause in fact and it must 
have been foreseeable.

\begin{enumerate}
    \item At the defendant's movie theater, somebody threw a bottle from a 
    balcony which struck and injured the plaintiff.
    \item The jury found the theater liable because it negligently failed to 
    remove ``rowdy persons'' from the balcony during the game. The appellate 
    court affirmed.
    \item The Texas Supreme Court clarified that proximate cause has two 
    elements: (1) cause in fact and (2) foreseeability. The court held that 
    the prosecution failed to show that the injuries would have occurred but 
    for the removal of the ``rowdy persons.'' Reversed.
\end{enumerate}

\subsubsection{Substantial Factor Test: \emph{Anderson v. Minneapolis, St. P. \& S. S. M. Ry. Co.}}

The substantial factor test creates liability for a defendant whose actions 
substantially caused the harm but were not the but-for cause.

\begin{enumerate}
    \item A spark from a railroad started a fire in a bog on one side of the 
    defendant's property. Another unrelated fire was burning on the other 
    side. The fire from the railroad destroyed the defendant's property, and a 
    few days later it joined with the other fire to make one big fire which 
    destroyed the defendant's home.
    \item The railroad argued that it cannot be held liable because the 
    defendant's house would have been destroyed by the other fire anyway.
    \item The trial court refused to instruct the jury to follow a rule from 
    an earlier case, \emph{Cook}, which held that there is no liability when 
    two fires jointly destroy property. On this basis, the trial court found 
    for the plaintiff. The railroad requested a motion for judgment 
    notwithstanding the verdict, which was denied.
    \item On appeal, the Supreme Court of Minnesota held that the 
    trial court was correct in refusing to apply the \emph{Cook} rule and 
    found for the plaintiffs.
    \item \textbf{Substantial factor test}: if two independent fires join to 
    cause property damage, there is joint liability, even if neither alone is 
    a but-for cause. Redundant causation is not necessary.
    \item Courts are split on whether to use the substantial factor test when 
    only one actor is liable. California courts do use it (and reject the 
    but-for test).
\end{enumerate}

\subsubsection{Substantial Factor Test II: \emph{Northington v. Marin}}

\begin{enumerate}
    \item The plaintiff, a prison inmate, sued the defendant, a prison guard, 
    for circulating rumors that labeled him a snitch and caused other inmates 
    to assault him. Other guards had spread the same rumors.
    \item The trial court found that although the defendant's action was not a 
    but-for cause (since the harm would have occurred without his action), his 
    contribution to the harm was nonetheless a \textbf{substantial factor}.
    \item The Tenth Circuit affirmed: ``Multiple tortfeasors who concurrently 
    cause an indivisible injury are jointly and severally liable; each can be 
    held liable for the entire injury.''
\end{enumerate}

\subsubsection{Proof of Harm: \emph{Herskovitz v. Group Health Cooperative of Puget Sound}}

Doctors can be held liable for negligently decreasing a patient's life 
expectancy even if the patient ``probably'' would not have lived (i.e., the 
patient's chance of survival was less than 50\%).

\begin{enumerate}
    \item The plaintiff brought the action on behalf of her husband, a 
    deceased lung cancer patient, against a doctor that negligently failed to 
    diagnose the patient's lung cancer on his first visit, proximately causing 
    his chance of survival to drop from 39 percent to 25 percent. Neither fact 
    was in dispute.
    \item The defendant argued that the plaintiff must prove that 
    the patient ``probably'' would have lived but for the negligence---that 
    is, without the doctor's negligence, the patient's chance of survival must 
    have been more than 50 percent.
    \item The trial court granted summary judgment 
    for the defendant on this argument.
    \item The Supreme Court of Washington 
    reversed, arguing that any other decision would mean a ``blanket release'' 
    for doctors' negligence any time the patient's chance of survival was less 
    than 50 percent. The court reasoned that if a defendant's acts have 
    \emph{increased the risk} of harm to the plaintiff, a jury should decide 
    whether the increased risk actually caused the harm in question.
\end{enumerate}

\subsubsection{Alternate Liability: \emph{Summers v. Tice}}

In cases where there are a small number of defendants, only one of them 
committed the harm, and we don't know which one, \textbf{each defendant is 
alternately liable for the full harm}. The burden of proof shifts to the 
defendant.

\begin{enumerate}
    \item The plaintiff and the two defendants were hunting quail. The two 
    defendants shot at a quail in the direction of the plaintiff. The 
    plaintiff suffered injuries, but it's not clear which defendant's shot was 
    the cause.
    \item The court held that the burden of proof shifts to the defendants to determine which one of them caused the injury. If 
    they cannot, ``each defendant is liable for the whole damage whether they 
    are deemed to be acting in concert or independently.''\footnote{Another 
    case with joint tortfeasors: \emph{Drabek v. Sabley} above (kids throwing 
    snowballs at cars)}
\end{enumerate}

\subsubsection{Market Share Liability: \emph{Sindell v. Abbott Laboratories}}

In California, when a product injurs a plaintiff but the specific manufacturer 
that created the harmful product cannot be identified, \emph{all} 
manufacturers of that product can be held liable according to their share of 
the market for that product.

\begin{enumerate}
    \item The plaintiff was harmed by DES, a prenatal drug intended to protect 
    against miscarriages but which turned out to pose significant danger to 
    unborn children. The plaintiff did not know which company manufactured the 
    specific drug her mother took, but since several companies manufactured 
    the drug according to the same formula, she sued them all.
    \item The companies won a dismissal at trial on the grounds that the 
    plaintiff could not identify which company caused the harm.
    \item The Supreme Court of California considered four theories of 
    liability:
    \begin{enumerate}
        \item The \emph{Summers} test: this failed because there are so many 
        defendants (over 200) that it is highly unlikely that any one of them 
        caused this specific injury.
        \item The ``concert of action'' theory: if the defendants had acted in 
        concert to cause the injury, they would be equally liable. In this 
        case, there was not sufficient evidence to show that the defendants 
        had a common plan to cause harm (e.g., by conducting inadequate safety 
        tests or giving insufficient safety warnings).
        \item ``Industry-wide'' or ``enterprise'' liability: if an entire 
        industry cooperates on an element of the harm in question---e.g., by 
        delegating safety testing to a trade association---they can be held 
        jointly liable. Here, the fact that DES manufacturers shared testing 
        and promotion methods did not establish industry-wide liability, 
        because (1) there are so many manufacturers and (2) safety standards 
        are mostly regulated by the FDA.
        \item \textbf{Market share liability}---a variation of the 
        \emph{Summers} test: each manufacturer's liability and share of the 
        damages are proportionate to its market share.
    \end{enumerate}
    \item Relying on the fourth theory, the court reversed, allowing the 
    plaintiff to proceed with her cause of action.
    \item Most states have not adopted market share liability.
    \item Defendants are allowed prove definitively that they did not 
    contribute to the harm (e.g., if they can show that they did not produce 
    the drug at the time).
    \item Some states require defendants to be joined so that a significant 
    share of the market is represented, and that missing market share 
    proportionally reduces the plaintiff's compensation. Usually (but not 
    always) this must be the nationwide market.\footnote{Casebook p. 229 n. 
    2.}
\end{enumerate}

\subsubsection{Toxic Exposure and Uncertain Harm: \emph{Ayers v. Township of Jackson}}

Hypothetical injuries from toxic exposure do not impose liability, but courts 
can award damages for relevant medical monitoring.

\begin{enumerate}
    \item A town in New Jersey was found to have caused toxic exposure by its 
    ``palpably unreasonable'' management of a landfill. The plaintiffs did not 
    develop any illnesses, but they sought to recover (1) damages for the 
    enhanced risk of future illness due to exposure and (2) regular medical 
    testing for diseases from exposure.
    \item The Supreme Court of New Jersey found 
    that the task of litigating hypothetical injuries would unreasonably strain 
    the tort system (although it suggests that the state legislature could 
    pass a remedy that allowed damages if toxic exposure caused a 
    ``statistically significant incidence of disease''). On the second claim, 
    it held for the plaintiffs.
\end{enumerate}

