\subsection{Overview}

\begin{enumerate}
    \item ``The failure to exercise the standard of care that a 
    \textbf{reasonably prudent [careful] person} would have exercised in a 
    similar situation.''\footnote{Black's Law.}
    \item Negligence requires proof that the defendant acted unreasonably.
    \item The standard of care is objective.
    \item Negligence has five factors:
    \begin{enumerate}
        \item Duty.
        \item Breach of duty.
        \begin{enumerate}
            \item Negligence.
            \item Breach of the standard of care.
            \item Failure to act as a reasonably careful person would under 
            the circumstances.
        \end{enumerate}
        \item Cause-in-fact.
        \item Proximate cause.
        \item Damages.
    \end{enumerate}
    \item Tort law generally doesn't lower the standard of care for people who 
    are unable to meet it.
    \item The Hand Formula is Judge Learned Hand's test for determining 
    negligence. It works best in scenarios where the actor takes a calculated 
    risk (e.g., business decisions). It is less useful in cases where the 
    actor was simply not paying attention.
    \begin{enumerate}
        \item B = Burden of precautions necessary to prevent an accident.
        \item P = Probability that an accident will occur.
        \item L = Magnitude of the loss if the accident occurs.
        \item Negligence exists if B\textless PL---i.e., if the burden of 
        precautions is less than the harm multiplied by the probability of 
        occurrence.
    \end{enumerate}
    \item Prosser: compare the utility of the risk with the gravity of the 
    loss.
\end{enumerate}

\subsubsection{\emph{Pitre v. Employers Liability Assurance Corporation}}

\begin{enumerate}
    \item The plaintiffs' son died when a patron at a carnival game was 
    winding up a pitch and hit him in the head. The trial court found in favor 
    of the plaintiffs. The determining factor, the Court of Appeal reasoned, 
    is how a ``reasonably prudent individual'' would have acted or what 
    precautions he would have taken under similar circumstances. 
    \textbf{Negligence occurs only if the danger is both foreseeable and 
    unreasonable.} The court held that the danger was foreseeable but not 
    unreasonable, and therefore there was no negligence.
\end{enumerate}

\subsubsection{\emph{United States Fidelity \& Guaranty Company v. Plovidba}}

\begin{enumerate}
    \item Inside a dark room, the hatch to a cargo hold on a ship was left 
    open. A longshoreman fell through it and died. The trial court found for 
    the defendant. Here, Richard Posner writing for the Seventh Circit applies 
    the Hand Formula, reasoning that B was relatively small (it would have 
    been easy to close the hatch or leave a light turned on) and L was high 
    (the victim died). P, however, was very small. There was no reason for the 
    longshoreman to enter the hold. In fact, he was probably there to steal 
    liquor, and the evidence suggests he knew the hatch was open and tried to 
    skirt around it. The shipowner was therefore not liable for negligence.
\end{enumerate}

\subsection{Standard of Conduct}

\subsubsection{\emph{Cordas v. Peerless Transp. Co.}}

\begin{enumerate}
    \item In emergencies, people are held to a lower standard.
    \item A man was mugged at gunpoint by two other men in New York City. He 
    chased after them. One of the muggers jumped into a taxi, held the driver 
    at gunpoint, and told him to drive. While the cab was in motion, the 
    driver jumped out, and a few seconds later, so did the hijacker. The cab 
    crashed into a sidewalk and injured the defendants. The trial court held 
    that the driver was not negligent because he acted as a reasonable person 
    would act under similar circumstances.
    \item Courts are divided on the question of whether juries should receive 
    special instructions regarding negligence claims in emergency 
    circumstances. On the one hand, it is redundant to reiterate that a 
    defendant must be held to the standard of what a reasonable person would 
    do in a similar emergency situation. Others claim it helps clarify the 
    standard.
    \item \textbf{Conditional privilege}: choose the lesser of two harms.
\end{enumerate}

\subsubsection{\emph{Breunig v. American Family Insurance Company}}

\begin{enumerate}
    \item A schizophrenic woman had a psychotic episode while driving her car. 
    The question was whether she had foreknowledge of her susceptibility to 
    such attacks. The general rule is that insanity or another mental 
    deficiency does not limit liability for negligence. In other words, 
    \textbf{insane people are held to the reasonable person standard}. The 
    court here notes that may be too harsh to exclude the insanity defense 
    when a driver is suddenly overcome without warning. The Supreme Court 
    agrees with the lower courts that the defendant did have the necessary 
    foreknowledge, and held for the plaintiff.  \item Two frameworks for 
    assessing liability from the sudden onset of mental illness:
    \begin{enumerate}
        \item Fairness: it's not fair to punish someone who could not have 
        avoided having a seizure.
        \item Loss distribution: if someone has to bear the cost of repairing 
        the harm, it should be the perpetrator, not the victim.
    \end{enumerate}
    \item One view: insanity constitutes a defense if there was no warning.
    \item Majority view (including CA): insanity does not create any defense 
    as to compensatory damages. Physical ailments, however, are taken into 
    account.
\end{enumerate}

\subsubsection{Child Standard and Adult Activity: \emph{Neumann v. Shlansky}}

\begin{enumerate}
    \item Should courts hold underage defendants to the standard of reasonable 
    adults?
    \item An eleven-year-old hit a golf ball that struck the defendant in the 
    knee, causing serious injury. Generally, children are held to the standard 
    of a \textbf{reasonable person of like age, intelligence, and experience 
    under the circumstances.} In this case, however, the child was engaging in 
    an ``adult activity,'' and therefore the court held him to the adult 
    reasonable person standard.
    \item Some states are moving from ``adult activity'' to ``inherently 
    dangerous activity.''
\end{enumerate}

\subsubsection{Professional Standards: \emph{Melville v. Southward}}

\begin{enumerate}
    \item The defendant, a podiatrist, operated on the plaintiff's foot. The 
    plaintiff sued for malpractice, and introduced the testimony of an 
    orthopedist, who questioned the necessity and sanitation of the operation. 
    The question is whether the orthopedist, a practitioner from a different 
    school of medicine, should have been allowed to testify about the standard 
    of care in podiatry. The trial court allowed the orthopedist to testify. 
    The Supreme Court of Colorado here agreed with the appellate court that 
    the testimony should not have been allowed because it was ``nothing more 
    than an expression of opinion that that the general practice of podiatry 
    did not meet the standard of care observed by an orthopedic surgeon.'' It 
    remanded the case for a new trial.
    \item In malpractice cases, the ``competent professional'' standard 
    replaces the ``reasonable person'' standard.
    \item There is disagreement about whether doctors in rural areas should be 
    held to different standards than urban doctors.
    \item Medical specialists in the same geographic region are often 
    reluctant to testify against each other---a ``conspiracy of silence.''
    \item In a limited range of cases, a jury of laypeople can determine 
    whether a practice met an acceptable standard of care.
    \item Why let a profession set its own standards?
\end{enumerate}

\subsubsection{\emph{Cobbs v. Grant}}

\begin{enumerate}
    \item Doctors are required to obtain \textbf{informed consent} from 
    patients. The plaintiff here sued a doctor who operated on a stomach ulcer 
    but did not discuss the surgery's inherent risks. Complications developed, 
    another operation was required, more complications developed, and so on.
    \item The plaintiff argued that (1) the doctor acted negligently in the 
    performance of the surgery (which the jury found in favor of the 
    plaintiff) and (2) that the doctor failed to obtain informed consent.
    \item The Supreme Court of California here noted that courts are divided 
    as to whether this type of tort should be deemed a \textbf{battery or 
    negligence}. The court aligned itself with a ``majority trend'' that 
    advocates reserving battery for cases where a doctor performs an operation 
    without the patient's consent. Generally, physicians are required to tell 
    patients about major risks (but not every minor risk) and obtain the 
    patient's consent. In this case, the court finds that there is not enough 
    evidence to show that the doctor acted negligently, and the case is 
    remanded for a new trial.
    \item \textbf{Failure to obtain informed consent can subject a physician 
    to negligence liability.} Unless the physician misrepresents the entire 
    procedure, most courts will not characterize the behavior as intentional 
    battery.
\end{enumerate}
