\subsection{Rules of Law and Negligence Per Se}

\subsubsection{Rules of Law}

\begin{enumerate}
    \item Juries are typically responsible for determining what constitutes 
    reasonable conduct under the circumstances. Judges, however, will 
    sometimes establish a \textbf{rule of law} for what constitutes negligent 
    conduct under particular circumstances.
    \item For instance, in \emph{Baltimore \& Ohio R.R.}, Justice Holmes 
    established a rule requiring drivers to get out of their cars and examine 
    railroad crossings (``stop, look, and listen'').
    \item Most courts do not use this approach because it's premised on 
    repetition of fact patterns. Fact patterns are rarely identical. Judges 
    may also not make the best rules (e.g., Holmes).
\end{enumerate}

\subsubsection{Rule of Law for Baseball Backstops: \emph{Akins v. Glens Falls 
City School District}}

\begin{enumerate}
    \item A foul ball injured a spectator at a baseball game. She sued the 
    ballpark's owners, the local school district, for negligence.
    \item The trial court held in favor of the plaintiff.
    \item The appellate court reversed, finding that the school district had 
    not acted negligently, and establishing a rule of law for ballpark 
    backstops.
    \item The dissent argued that such a rule ``robs the jury'' of the ability 
    to consider important circumstances and locks the law in ``a position that 
    is certain to become outdated.''
\end{enumerate}

\subsubsection{Negligence Per Se}
% replace ยง with \S\ throughout
\begin{enumerate}
    \item Under negligence per se, \textbf{liability exists when the defendant 
    violates a statute}.
    \item \textbf{Presumption of negligence}: in some states, a jury 
    \textbf{must presume} negligence when a statute is breached. The defendant 
    is free to rebut. California follows this rule, with a few 
    exceptions.\footnote{Cal. Evid. C. \S\ 669.  See course reader p. 11.} 
    \item In other states, juries are free to (but need not) \textbf{infer} 
    that breach of statute constitutes negligence---e.g., a car doesn't slow 
    down and hits a pedestrian in a crosswalk.
    \item Plaintiff can, and usually will, plead both common law negligence 
    and negligence per se.
    \item Compliance with a statute is generally not proof of due care.
    \item \textbf{Statutory purpose doctrine}: for the statute to be relevant, the 
    harm that occurred must have been the type that the statute was intended 
    to prevent. However, statutory purpose can sometimes be unclear, and it 
    may change through time.
    \item Dual purpose doctrine: a statute may have more than one narrow 
    purpose.
    \item Generally (including Levy): proof of compliance with a statute is 
    \textbf{never} proof of due care. Criminal statutes set a minimum of 
    conduct that could be below what we'd call due care. Some cases take the 
    opposing view.
    \item A federal statute may preempt what would otherwise be a state cause 
    of action. Types of preemption: explicit, conflict, and field.
\end{enumerate}

\subsubsection{Statutory Purpose Doctrine: \emph{Wawanesa Mutual Insurance Co. v. Matlock}}

To apply negligence per se, the harm in question must be the type of harm the 
statute aims to prevent.

\begin{enumerate}
    \item A minor bought cigarettes for another minor, who later dropped the 
    cigarette and caused a fire that led to significant property damage. The 
    insurer sued the first minor's father.
    \item The trial court found for the insurer.
    \item The appellate court reversed, holding that that the statute in 
    question (preventing people from purchasing tobacco for minors) was meant 
    to protect against the health hazards of tobacco, not the fire hazard, and 
    therefore cannot be used to establish a standard of conduct in this case.
\end{enumerate}

\subsubsection{Interpreting Legislative Intent: \emph{Stachniewicz v. Mar-Cam 
Corporation}}

Negligence per se applies if the legislature intended to prevent the harm in 
question.

\begin{enumerate}
    \item A patron injured in bar brawl sued the bar owner. The plaintiff 
    relied on (1) an Oregon statute which prohibits giving alcohol to an 
    intoxicated person and (2) an Oregon regulation that prevents bar owners 
    from allowing disorderly conduct on their premised.
    \item The trial court found for the defendant. 
    \item The appellate court overturned, reasoning that (1) the 
    statute was inapplicable because the brawler was already drunk when he 
    arrived, so there was no way to tell if another drink caused the brawl, but 
    (2) the regulation was intended specifically to protect customers from 
    injury, and therefore can be an appropriate standard for negligence in 
    this case.
\end{enumerate}

\subsubsection{Statutory Purpose Doctrine II: \emph{Gorris v. Scott}}

See \emph{Wawanesa}, above.

\begin{enumerate}
    \item Several sheep on a ship were swept overboard. The plaintiff sued the 
    shipowner, arguing that the Contagious Diseases (Animals) Act required the 
    shipowner to enclose the sheep in pens of certain dimensions, which the 
    shipowner failed to do.
    \item The court found in favor of the shipowner, 
    reasoning that the Act was intended to prevent the spread of contagious 
    diseases, not to prevent sheep from falling overboard.
\end{enumerate}
